---
title: "Deterministic and Stochastic Epidemic Modelling in R"
author: "Will Lodder"
date: last-modified
format:
  html:
    toc: true
    code-fold: true
  pdf:
    toc: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

library(ggplot2)
library(deSolve)

source("R-code/single-Euler-step.r")
source("R-code/daily-change-in-cases.r")
source("R-code/model-error-function.r")
source("R-code/beta-param-fitting.r")
source("R-code/varying-beta-values.r")
source("R-code/exact-SIRV-soln.r")
source("R-code/one-step-Markov.r")
source("R-code/beta-param-fitting-Markov.r")
source("R-code/Markov-simulation.r")

omicron_data <- read.csv("data/omicron.csv")
omicron_data$date <- as.Date(omicron_data$date)

N_UK <- 56490048
```




The SIR model



```{r}
print(head(simulate_sir_euler(0.6, 0.2, 0.9, 0.1, 40, 0.1)))
```






```{r}

beta <- 0.6
gamma <- 0.2
x0 <- 0
y0 <- 1
T <- 40
h <- 0.1


zero_susceptibles <- simulate_sir_euler(0.6, 0.2, 0, 1, 40, 0.1)

#exact soln
zero_susceptibles$exact <- y0 * exp(-gamma * zero_susceptibles$t)

#defining absolute value of differences
differences <- abs(zero_susceptibles$y - zero_susceptibles$exact)

p0 <- ggplot(zero_susceptibles, aes(x = t, y = differences)) +
    geom_line() +
    geom_point(size = 0.8) +
    xlab("Time") +
    ylab("Absolute difference")

print(p0)
```





```{r}


#now want to use h = 7 days
#time now moves in weekly steps, will use t = 42 as its the closest 
#multiple of 7 to 40 and prevents issue 
#of fractional step sizes

weekly_estimate <- simulate_sir_euler(0.6, 0.2, 0.9, 0.1, 42, 7)

p1 <- ggplot(weekly_estimate, aes(x = t)) +
    geom_line(aes(y = x, colour = "Susceptible")) +
    geom_line(aes(y = y, colour = "Infected")) +
    #inserting legend as we have multiple lines
    scale_colour_manual(
        name = "Variable",
        values = c("Susceptible" = "blue", "Infected" = "red")
    ) +
    xlab("Time") +
    ylab("Population Proportion")

print(p1)


```

As we can see, using a larger time step such as 7 days can cause issues with the estimation. We observe population proportions greater than 1, which is clearly impossible. This therefore emphasizes we should use smaller step sizes, sich as $h = 0.1$ to increase the accuracy of the euler approximation. 







```{r}


rows_omicron_data <- nrow(omicron_data)
initial_population <- 56490048

#just defining time vector, will adjoin this to df later
omicron_data$time <- 0:(rows_omicron_data - 1)

#defining the incidence as in the question
omicron_data$incidence <- omicron_data$cases / initial_population

#adding the vectors to our dataframe
p2 <- ggplot(omicron_data, aes(x = time, y = incidence)) +
    geom_line(colour = "#ff00ff") +
    xlab("Time") +
    ylab("Incidence")

print(p2)
```




```{r}
print(head(calculate_infectious_change(0.6, 0.2, 0.9, 0.1, 40, 0.1)))
```



```{r}
beta_variable_function <- function_factory_beta(0.2, 0.9, 0.1, 40, 0.1)

beta_0_5 <- beta_variable_function(0.5)
beta_0_6 <- beta_variable_function(0.6)
beta_0_7 <- beta_variable_function(0.7)

#making a data frame which i will plot
beta_variable_df <- data.frame(
    t = beta_0_5$t,
    beta_0_5 = beta_0_5$incidence,
    beta_0_6 = beta_0_6$incidence,
    beta_0_7 = beta_0_7$incidence
)

p3 <- ggplot(beta_variable_df, aes(x = t)) +
    geom_line(aes(y = beta_0_5, colour = "beta = 0.5")) +
    geom_line(aes(y = beta_0_6, colour = "beta = 0.6")) +
    geom_line(aes(y = beta_0_7, colour = "beta = 0.7")) +
    scale_colour_manual(
        name = "Model",

        values = c(
            "beta = 0.5" = "blue",
            "beta = 0.6" = "red",
            "beta = 0.7" = "green"
        )
    ) +
    xlab("Time") +
    ylab("Incidence")

print(p3)
```









```{r}

print(calculate_incidence_sse(0.6, 0.1, 0.9, 0.1, omicron_data))

```



```{r}

error_function <- factory_error(0.1, 1 - 0.004, 0.004, omicron_data)

#betas as outlined in question, storing initial vector of zeros for the errors
beta_variable <- seq(from = 0.1, to = 0.2, by = 0.01)
errors_variable <- rep(0, length(beta_variable))


#updating our errors vector
for (i in seq_along(beta_variable)){
    errors_variable[i] <- error_function(beta_variable[i])
}


#storing data frame which will be plotted
df_p4 <- data.frame(
    betas = beta_variable,
    errors = errors_variable
)


p4 <- ggplot(df_p4, aes(x = betas, y = errors)) +
    geom_line(colour = "purple") + 
    xlab("Beta") +
    ylab("Squared difference")

print(p4)
```




```{r}




#now using it for given data
findmin_function <- findmin_factory(0.1, 1 - 0.004, 0.004, omicron_data)
best_beta <- findmin_function(0.1, 0.2, 100)

print(best_beta)
```




```{r}
#incidence
observed_incidence <- omicron_data$cases / initial_population

model_incidence <- calculate_infectious_change(best_beta, 0.1, 1 - 0.004, 0.004, 46, 1)



df_p5 <- data.frame(
    t = model_incidence$t,
    observed_incidence = observed_incidence,
    model_incidence = model_incidence$incidence
)




p5 <- ggplot(df_p5, aes(x = t)) +
    geom_line(aes(y = observed_incidence, colour = "Observed")) +
    geom_line(aes(y = model_incidence, colour = "Model")) +
    scale_colour_manual(
        name = "Variable",
        values = c("Observed" = "red", "Model" = "blue")
    ) +
    xlab("Time") +
    ylab("Incidence")


print(p5)
```





```{r}



euler_p6 <- simulate_sir_euler(0.15, 0.1, 1 - 0.004, 0.004, 40, 1)

#converting to dataframe so its plottable
SIRV_soln_p6 <- solve_sirv_ode(0.15, 0.1, 0, 1 - 0.004, 0.004, 0, 40, 1)
SIRV_soln_p6 <- as.data.frame(SIRV_soln_p6)


x_diffs <- abs(euler_p6$x - SIRV_soln_p6$x)
y_diffs <- abs(euler_p6$y - SIRV_soln_p6$y)
times_p6 <- SIRV_soln_p6$time


df_p6 <- data.frame(
    x_diffs = x_diffs,
    y_diffs = y_diffs, 
    time = times_p6
)



#plotting difference between the SIRV models solution for paramaters
#and the solution given by euler approx. for each parameter

p6 <- ggplot(df_p6, aes(x = time)) +
    geom_line(aes(y = x_diffs, colour = "x differences")) +
    geom_line(aes(y = y_diffs, colour = "y differences")) +
    scale_colour_manual(
        name = "Variable",
        values = c(
            "x differences" = "blue",
            "y differences" = "red"
        )
    ) +
    xlab("Time") +
    ylab("Absolute difference")


print(p6)
```





```{r}

#following standard as before
solution_p7 <- solve_sirv_ode(0.15, 0.1, 0.002, 1 - 0.004, 0.004, 0, 200, 1)


solution_p7 <- as.data.frame(solution_p7)

#pulling values from solution_p7
time_p7 <- solution_p7$time
x_p7 <- solution_p7$x
y_p7 <- solution_p7$y
v_p7 <- solution_p7$v

#getting remaining population
remainder <- 1 - solution_p7$x - solution_p7$y - solution_p7$v


#making the dataframe
df_p7 <- data.frame(
    x_p7 = x_p7,
    y_p7 = y_p7,
    v_p7 = v_p7,
    removed_proportion = remainder,
    time_p7 = time_p7
)

#plotting
p7 <- ggplot(df_p7, aes(x = time_p7)) +
    geom_line(aes(y = x_p7, colour = "susceptible proportion")) +
    geom_line(aes(y = y_p7, colour = "infectious proportion")) +
    geom_line(aes(y = v_p7, colour = "vaccinated proportion")) +
    geom_line(aes(y = remainder, colour = "removed proportion")) +
    scale_colour_manual(
        name = "Variable",
        
        values = c(
            "susceptible proportion" = "blue",
            "infectious proportion" = "red", 
            "vaccinated proportion" = "green", 
            "removed proportion" = "purple"
        )
    ) +
    xlab("Time") +
    ylab("Variable values")
    

print(p7)
```





```{r}


mu_sequence <- seq(
  from = 0,
  to = 0.005,
  by = 0.001
)

df_list <- vector(
  mode = "list",
  length = length(mu_sequence)
)

for (i in seq_along(mu_sequence)) {
  sol <- solve_sirv_ode(
    beta = 0.15,
    gamma = 0.1,
    mu = mu_sequence[i],
    x0 = 1 - 0.004,
    y0 = 0.004,
    v0 = 0,
    T = 200,
    h = 1
  )

  sol <- as.data.frame(sol)

  df_list[[i]] <- data.frame(
    time = sol$time,
    infectious_proportion = sol$y,
    mu = factor(
      mu_sequence[i],
      levels = mu_sequence
    )
  )
}

df_p8 <- do.call(
  rbind,
  df_list
)

p8 <- ggplot(
  df_p8,
  aes(
    x = time,
    y = infectious_proportion,
    colour = mu
  )
) +
  geom_line() +
  labs(
    x = "Time",
    y = "Infectious population proportion",
    colour = expression(mu),
    title = "Effect of the vaccination rate on infectious prevalence"
  )

print(p8)







```




As we can see, if there exists no vaccination against the illness over time, we get a larger peak of daily incidence. This makes sense logically, as with no vaccination, a greater proportion of the population will be within the susceptible population, therefore allowing for more people in total to be infected. Further, with higher vaccination rates we observe a smaller peak in incidence. 



We will now implement a markov model to investigate how things may change. The markov model may be more useful in practice as we are able to compute confidence intervals.

```{r}

#print(head(omicron_data))
T <- 1000


```


We now create a function which performs a single markov step, and then define a funciton which iterates this over T steps of size 1. 
```{r}
#next step will be to implement the markov process and iterate over

#defining intital vectors as was in the project
df1 <- initial_populations(1-0.004, 0.004, 0, T = T)
print(head(df1))

```

defining an N_eff because this should display a little more stochastic variance. useful for visualisation.


```{r}

T <- 1000

#df1 <- simulate_markov(
  #beta = 0.15,
  #gamma = 0.1,
  #mu = 0.001,
  #x0 = 1 - 0.004,
  #y0 = 0.004,
  #v0 = 0,
  #T = T,
  #N = 56490048
#)

#print(head(df1))


#plot(df1$t, df1$incidence / 56490048, type = "l")




# Real UK population used for observed case proportions
N_UK <- 56490048

# Effective population size used inside stochastic Markov model
# Smaller N_eff = more visible stochastic variation
N_eff <- 50000

T <- 1000

set.seed(4)

df1 <- simulate_markov(
    
  beta = best_beta,
  gamma = 0.1,
  mu = 0.000,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  T = 46,
  N = N_eff
)

omicron_data$date <- as.Date(omicron_data$date)

# Remove initial model row because incidence[1] is artificial
model_incidence <- df1[-1, c("t", "incidence")]

n <- min(nrow(omicron_data), nrow(model_incidence))

comparison_df <- data.frame(
  date = omicron_data$date[1:n],
  observed_cases = omicron_data$cases[1:n],
  observed_prop = omicron_data$cases[1:n] / N_UK,
  model_incidence = model_incidence$incidence[1:n],
  model_prop = model_incidence$incidence[1:n] / N_eff
)

p1 <- ggplot(comparison_df, aes(x = date)) +
  geom_line(aes(y = observed_prop, linetype = "Observed daily cases / N")) +
  geom_line(aes(y = model_prop, linetype = "Markov model incidence / N_eff")) +
  labs(
    x = "Date",
    y = "Daily incidence proportion",
    title = "Observed Omicron Daily Cases vs Markov Model Incidence",
    linetype = "Series"
  ) +
  theme_minimal()




print(p1)


```


This appears way off. this is because we are using the value beta_best which calculates daily indicence as the number of new infected individuals - number of new recovered individuals, which is not used in practice. will now make a code block to fit a better parameter for beta for our markov model

```{r}



fit_markov <- optimize(
  f = markov_error_beta,
  interval = c(0.01, 0.5),
  data = omicron_data,
  gamma = 0.1,
  mu = 0,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  N_UK = N_UK
)

beta_best_markov <- fit_markov$minimum
markov_min_error <- fit_markov$objective

print(beta_best_markov)
print(markov_min_error)

```


We now use this corrected beta value and plot a single path using this. 
```{r}

df2 <- simulate_markov(
    
  beta = beta_best_markov,
  gamma = 0.1,
  mu = 0.000,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  T = 46,
  N = N_eff
)

omicron_data$date <- as.Date(omicron_data$date)

# Remove initial model row because incidence[1] is artificial
model_incidence_2 <- df2[-1, c("t", "incidence")]

n <- min(nrow(omicron_data), nrow(model_incidence_2))

comparison_df_2 <- data.frame(
  date = omicron_data$date[1:n],
  observed_cases = omicron_data$cases[1:n],
  observed_prop = omicron_data$cases[1:n] / N_UK,
  model_incidence = model_incidence_2$incidence[1:n],
  model_prop = model_incidence_2$incidence[1:n] / N_eff
)

p2 <- ggplot(comparison_df_2, aes(x = date)) +
  geom_line(aes(y = observed_prop, linetype = "Observed daily cases / N")) +
  geom_line(aes(y = model_prop, linetype = "Markov model incidence / N_eff")) +
  labs(
    x = "Date",
    y = "Daily incidence proportion",
    title = "Observed Omicron Daily Cases vs Markov Model Incidence",
    linetype = "Series"
  ) +
  theme_minimal()




print(p2)



```



We now want to find confidence bands, by taking M simulations and taking quantiles. 

```{r}

simulation_summary <- do.call(
  rbind,
  lapply(split(many_sims$incidence_prop, many_sims$t), function(z) {
    data.frame(
      mean = mean(z),
      lower = as.numeric(quantile(z, 0.025)),
      upper = as.numeric(quantile(z, 0.975))
    )
  })
)



many_sims <- simulate_many_markov(
M = M,
beta = beta_best_markov,
gamma = 0.1,
mu = 0,
x0 = 1 - 0.004,
y0 = 0.004,
v0 = 0,
T = T_fit,
N_eff = N_eff
)




simulation_summary$t <- as.numeric(rownames(simulation_summary))
rownames(simulation_summary) <- NULL

# Reorder columns nicely
simulation_summary <- simulation_summary[, c("t", "mean", "lower", "upper")]

# Add dates and observed incidence
n <- min(nrow(omicron_data), nrow(simulation_summary))

plot_df <- data.frame(
  date = as.Date(omicron_data$date[1:n]),
  observed_prop = omicron_data$cases[1:n] / N_UK,
  mean = simulation_summary$mean[1:n],
  lower = simulation_summary$lower[1:n],
  upper = simulation_summary$upper[1:n]
)


p_many <- ggplot(plot_df, aes(x = date)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25) +
  geom_line(aes(y = mean, linetype = "Mean Markov incidence")) +
  geom_line(aes(y = observed_prop, linetype = "Observed daily cases / N")) +
  labs(
    x = "Date",
    y = "Daily incidence proportion",
    title = "Observed Omicron Cases vs Markov Simulation Envelope",
    linetype = "Series"
  ) +
  theme_minimal()

print(p_many)



```



