---
title: "Epidemic Modelling in R"
author: "Will Lodder"
date: "`r Sys.Date()`"
format: pdf
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




The SIR model



```{r}


library(ggplot2)



euler <- function(beta, gamma, x0, y0, T, h) {
    #defining our number of steps
    n <- as.integer(T / h)

  #checking h divides T nicely
  #stops errors that could arise if we have fractional step sizes

    if (!isTRUE(all.equal(n * h, T))) {
        stop("h doesnt divide T exactly")
    }


    #defining some initial vectors whcih we will replace later
    t <- seq(0, T, by = h)
    x <- rep(0, n + 1)
    y <- rep(0, n + 1)



    #making sure initial values coincide with question statement
    x[1] <- x0
    y[1] <- y0


    #partial differential eqtns given in question
    x_prime <- function(x_value, y_value) {
        -beta * x_value * y_value
    }

    y_prime <- function(x_value, y_value) {
        beta * x_value * y_value - gamma * y_value
    }


    #updating our vectors for each i
    for (i in 1:n) {
        x[i + 1] <- x[i] + h * x_prime(x[i], y[i])
        y[i + 1] <- y[i] + h * y_prime(x[i], y[i])
    }

    data.frame(t = t, x = x, y = y)
}

print(head(euler(0.6, 0.2, 0.9, 0.1, 40, 0.1)))
```






```{r}

beta <- 0.6
gamma <- 0.2
x0 <- 0
y0 <- 1
T <- 40
h <- 0.1


zero_susceptibles <- euler(0.6, 0.2, 0, 1, 40, 0.1)

#exact soln
zero_susceptibles$exact <- y0 * exp(-gamma * zero_susceptibles$t)

#defining absolute value of differences
differences <- abs(zero_susceptibles$y - zero_susceptibles$exact)

p0 <- ggplot(zero_susceptibles, aes(x = t, y = differences)) +
    geom_line() +
    geom_point(size = 0.8) +
    xlab("Time") +
    ylab("Absolute difference")

print(p0)
```





```{r}


#now want to use h = 7 days
#time now moves in weekly steps, will use t = 42 as its the closest 
#multiple of 7 to 40 and prevents issue 
#of fractional step sizes

weekly_estimate <- euler(0.6, 0.2, 0.9, 0.1, 42, 7)

p1 <- ggplot(weekly_estimate, aes(x = t)) +
    geom_line(aes(y = x, colour = "Susceptible")) +
    geom_line(aes(y = y, colour = "Infected")) +
    #inserting legend as we have multiple lines
    scale_colour_manual(
        name = "Variable",
        values = c("Susceptible" = "blue", "Infected" = "red")
    ) +
    xlab("Time") +
    ylab("Population Proportion")

print(p1)


```

As we can see, using a larger time step such as 7 days can cause issues with the estimation. We observe population proportions greater than 1, which is clearly impossible. This therefore emphasizes we should use smaller step sizes, sich as $h = 0.1$ to increase the accuracy of the euler approximation. 







```{r}

omicron_data <- read.csv("omicron.csv")


rows_omicron_data <- nrow(omicron_data)
initial_population <- 56490048

#just defining time vector, will adjoin this to df later
omicron_data$time <- 0:(rows_omicron_data - 1)

#defining the incidence as in the question
omicron_data$incidence <- omicron_data$cases / initial_population

#adding the vectors to our dataframe
p2 <- ggplot(omicron_data, aes(x = time, y = incidence)) +
    geom_line(colour = "#ff00ff") +
    xlab("Time") +
    ylab("Incidence")

print(p2)
```




```{r}

#incidence formula given in question as follows:
# y(kh) - y((k-1)h) for k in \mathbb{N}
incidence <- function(beta, gamma, x0, y0, T, h) {
    model_output <- euler(beta, gamma, x0, y0, T, h)

    #adding 0 to the start to make the vector lengths natch up to t vector
    incidence_values <- c(0, diff(model_output$y))

    data.frame(
        t = model_output$t,
        incidence = incidence_values
    )
}

print(head(incidence(0.6, 0.2, 0.9, 0.1, 40, 0.1)))
```



```{r}


#this returns the incidence function of beta
function_factory_2c <- function(gamma, x0, y0, T, h) {
    function(beta) {
        incidence(beta, gamma, x0, y0, T, h)
    }
}

beta_variable_function <- function_factory_2c(0.2, 0.9, 0.1, 40, 0.1)

beta_0_5 <- beta_variable_function(0.5)
beta_0_6 <- beta_variable_function(0.6)
beta_0_7 <- beta_variable_function(0.7)

#making a data frame which i will plot
beta_variable_df <- data.frame(
    t = beta_0_5$t,
    beta_0_5 = beta_0_5$incidence,
    beta_0_6 = beta_0_6$incidence,
    beta_0_7 = beta_0_7$incidence
)

p3 <- ggplot(beta_variable_df, aes(x = t)) +
    geom_line(aes(y = beta_0_5, colour = "beta = 0.5")) +
    geom_line(aes(y = beta_0_6, colour = "beta = 0.6")) +
    geom_line(aes(y = beta_0_7, colour = "beta = 0.7")) +
    scale_colour_manual(
        name = "Model",

        values = c(
            "beta = 0.5" = "blue",
            "beta = 0.6" = "red",
            "beta = 0.7" = "green"
        )
    ) +
    xlab("Time") +
    ylab("Incidence")

print(p3)
```









```{r}

error <- function(beta, gamma, x0, y0, data) {
    #largest time value in data
    T <- data$time[nrow(data)]
    #assuming equal step size
    h <- data$time[2] - data$time[1]

    #incidence from data set
    dt <- data$incidence
    #estimated incidence from previous question function
    mt <- incidence(beta, gamma, x0, y0, T, h)$incidence

    #checking lengths of mt and dt are same
    if (length(mt) != length(dt)) {
        stop("Model output and data incidence have different lengths.")
    }

    sum((mt - dt)^2)
}

print(error(0.6, 0.1, 0.9, 0.1, omicron_data))

```



```{r}
factory_error <- function(gamma, x0, y0, data) {
    output <- function(beta) {
        error(beta, gamma, x0, y0, data)
    }

    return(output)
}



error_function <- factory_error(0.1, 1 - 0.004, 0.004, omicron_data)

#betas as outlined in question, storing initial vector of zeros for the errors
beta_variable <- seq(from = 0.1, to = 0.2, by = 0.01)
errors_variable <- rep(0, length(beta_variable))


#updating our errors vector
for (i in seq_along(beta_variable)){
    errors_variable[i] <- error_function(beta_variable[i])
}


#storing data frame which will be plotted
df_p4 <- data.frame(
    betas = beta_variable,
    errors = errors_variable
)


p4 <- ggplot(df_p4, aes(x = betas, y = errors)) +
    geom_line(colour = "purple") + 
    xlab("Beta") +
    ylab("Squared difference")

print(p4)
```




```{r}


findmin_factory <- function(gamma, x0, y0, data){
    #we want to find minimum beta between beta_min, nbeta_max with N steps
    #we wish to return a function of beta_min, beta_max, N

    error_function <- factory_error(gamma, x0, y0, data)

    #will end up returning this function
    findmin <- function(beta_min, beta_max,N){

        errors_variable <- rep(0, N)

        #will find errors for each of these
        #then will return beta with lowest error
        beta_sequence <- seq(from = beta_min, to = beta_max, length.out = N)


        for (i in seq_along(beta_sequence)) {
            errors_variable[i] <- error_function(beta_sequence[i])
        }


        #finding where minimum beta value is, then returning it
        min_beta_position <- which.min(errors_variable)
        return(beta_sequence[min_beta_position])


    }
    
return(findmin)

}

#now using it how question specified
findmin_function <- findmin_factory(0.1, 1 - 0.004, 0.004, omicron_data)
best_beta <- findmin_function(0.1, 0.2, 100)

print(best_beta)
```




```{r}
#incidence
observed_incidence <- omicron_data$cases / initial_population

model_incidence <- incidence(best_beta, 0.1, 1 - 0.004, 0.004, 46, 1)



df_p5 <- data.frame(
    t = model_incidence$t,
    observed_incidence = observed_incidence,
    model_incidence = model_incidence$incidence
)




p5 <- ggplot(df_p5, aes(x = t)) +
    geom_line(aes(y = observed_incidence, colour = "Observed")) +
    geom_line(aes(y = model_incidence, colour = "Model")) +
    scale_colour_manual(
        name = "Variable",
        values = c("Observed" = "red", "Model" = "blue")
    ) +
    xlab("Time") +
    ylab("Incidence")


print(p5)
```





```{r}

library(deSolve)
#had not used DeSolve prior, used similar method as in:
#https://cran.r-project.org/web/packages/deSolve/vignettes/deSolve.pdf

solve <- function(beta, gamma, mu, x0, y0, v0, T, h){

    state <- c(x = x0, y = y0, v = v0)
    parameters <- c(beta = beta , gamma = gamma , mu = mu )

    SIRV <- function(t, state, parameters) {

        with(as.list(c(state, parameters)), {
        dx <-  -beta * x * y - mu * x
        dy <-  beta * x * y - gamma * y
        dv <- mu * x
        list(c(dx, dy, dv))
    })
}

tim <- seq(0, T, by = h)



sol <- ode(y = state, times = tim, func = SIRV, parms = parameters)


return(sol)

}



euler_p6 <- euler(0.15, 0.1, 1 - 0.004, 0.004, 40, 1)

#converting to dataframe so its plottable
SIRV_soln_p6 <- solve(0.15, 0.1, 0, 1 - 0.004, 0.004, 0, 40, 1)
SIRV_soln_p6 <- as.data.frame(SIRV_soln_p6)


x_diffs <- abs(euler_p6$x - SIRV_soln_p6$x)
y_diffs <- abs(euler_p6$y - SIRV_soln_p6$y)
times_p6 <- SIRV_soln_p6$time


df_p6 <- data.frame(
    x_diffs = x_diffs,
    y_diffs = y_diffs, 
    time = times_p6
)



#plotting difference between the SIRV models solution for paramaters
#and the solution given by euler approx. for each parameter

p6 <- ggplot(df_p6, aes(x = time)) +
    geom_line(aes(y = x_diffs, colour = "x differences")) +
    geom_line(aes(y = y_diffs, colour = "y differences")) +
    scale_colour_manual(
        name = "Variable",
        values = c(
            "x differences" = "blue",
            "y differences" = "red"
        )
    ) +
    xlab("Time") +
    ylab("Absolute difference")


print(p6)
```





```{r}

#following standard as before
solution_p7 <- solve(0.15, 0.1, 0.002, 1 - 0.004, 0.004, 0, 200, 1)


solution_p7 <- as.data.frame(solution_p7)

#pulling values from solution_p7
time_p7 <- solution_p7$time
x_p7 <- solution_p7$x
y_p7 <- solution_p7$y
v_p7 <- solution_p7$v

#getting remaining population
remainder <- 1 - solution_p7$x - solution_p7$y - solution_p7$v


#making the dataframe
df_p7 <- data.frame(
    x_p7 = x_p7,
    y_p7 = y_p7,
    v_p7 = v_p7,
    removed_proportion = remainder,
    time_p7 = time_p7
)

#plotting
p7 <- ggplot(df_p7, aes(x = time_p7)) +
    geom_line(aes(y = x_p7, colour = "susceptible proportion")) +
    geom_line(aes(y = y_p7, colour = "infectious proportion")) +
    geom_line(aes(y = v_p7, colour = "vaccinated proportion")) +
    geom_line(aes(y = remainder, colour = "removed proportion")) +
    scale_colour_manual(
        name = "Variable",
        
        values = c(
            "susceptible proportion" = "blue",
            "infectious proportion" = "red", 
            "vaccinated proportion" = "green", 
            "removed proportion" = "purple"
        )
    ) +
    xlab("Time") +
    ylab("Variable values")
    

print(p7)
```





```{r}


#making a function factory so i can just do a sequence of mu's
#should mean i dont have to define loads of different ones

factory_vaccinated_prop <- function(beta, gamma, x0, y0, v0, T, h) {
    output <- function(mu) {
        solve(beta, gamma, mu, x0, y0, v0, T, h)
    }

    return(output)
}



mu_function <- factory_vaccinated_prop(0.15, 0.1, 1 - 0.004, 0.004, 0, 200, 1)

#mu sequence 
mu_sequence <- seq(from = 0, to = 0.005, by = 0.001)



#making an empty vector of length 6, will put data frames in this list
df_list <- vector("list", length(mu_sequence))



for (i in seq_along(mu_sequence)) {

    #storing the solution as a data frame 
    #will then input into our df_list
    #will end up with a list of data frames
    sol <- as.data.frame(mu_function(mu_sequence[i]))

    df_list[[i]] <- data.frame(
        time = sol$time,
        y = sol$y,
        mu = paste0("mu = ", mu_sequence[i])
    )
}


#combining the df_list into a large singular data frame
#means i am able to plot it using ggplot
df_p8 <- do.call(rbind, df_list)


#plotting
p8 <- ggplot(df_p8, aes(x = time, y = y, colour = mu)) +
    geom_line() +
    xlab("Time") +
    ylab("Proportion of infection individuals")

print(p8)


```




As we can see, if there exists no vaccination against the illness over time, we get a larger peak of daily incidence. This makes sense logically, as with no vaccination, a greater proportion of the population will be within the susceptible population, therefore allowing for more people in total to be infected. Further, with higher vaccination rates we observe a smaller peak in incidence. 



We will now implement a markov model to investigate how things may change. The markov model may be more useful in practice as we are able to compute confidence intervals.

```{r}
omicron_data <- read.csv("/Users/willlodder/Desktop/Math245/omicron.csv")

#print(head(omicron_data))
T <- 1000


#making a function which sets up a data frame given our starting params
initial_populations <- function(x0, y0, v0, T, N = 56490048) {
  
  starting_susceptible <- as.integer(round(N * x0))
  starting_infectious  <- as.integer(round(N * y0))
  starting_vaccinated  <- as.integer(round(N * v0))
  starting_removed     <- N - starting_susceptible - starting_infectious - starting_vaccinated
  
  populations_df <- data.frame(
    t = 0:T,
    S = integer(T + 1),
    I = integer(T + 1),
    R = integer(T + 1),
    V = integer(T + 1),
    x = numeric(T + 1),
    y = numeric(T + 1),
    r = numeric(T + 1),
    v = numeric(T + 1),
    incidence = integer(T + 1)
  )
  
  populations_df$S[1] <- starting_susceptible
  populations_df$I[1] <- starting_infectious
  populations_df$R[1] <- starting_removed
  populations_df$V[1] <- starting_vaccinated
  
  populations_df$x[1] <- populations_df$S[1] / N
  populations_df$y[1] <- populations_df$I[1] / N
  populations_df$r[1] <- populations_df$R[1] / N
  populations_df$v[1] <- populations_df$V[1] / N
  
  populations_df$incidence[1] <- 0
  
  return(populations_df)
}


```


We now create a function which performs a single markov step, and then define a funciton which iterates this over T steps of size 1. 
```{r}
#next step will be to implement the markov process and iterate over


one_step_markov <- function(S, I, R, V, beta, gamma, mu, N) {
  
  # 1. Infection and vaccination hazards
  lambda_inf <- beta * I / N
  lambda_vac <- mu
  
  # 2. Total hazard out of S
  lambda_total <- lambda_inf + lambda_vac
  
  # 3. Convert hazards into one-day transition probabilities
  if (lambda_total > 0) {
    p_leave_S <- 1 - exp(-lambda_total)
    
    p_SI <- (lambda_inf / lambda_total) * p_leave_S
    p_SV <- (lambda_vac / lambda_total) * p_leave_S
  } else {
    p_SI <- 0
    p_SV <- 0
  }
  
  p_stay <- 1 - p_SI - p_SV
  
  # 4. Removal probability
  p_IR <- 1 - exp(-gamma)
  
  # 5. Random transitions from S
  S_transitions <- as.vector(
    rmultinom(
      n = 1,
      size = S,
      prob = c(p_SI, p_SV, p_stay)
    )
  )
  
  N_SI <- S_transitions[1]  # new infections
  N_SV <- S_transitions[2]  # new vaccinations
  
  # 6. Random recoveries from I
  N_IR <- rbinom(
    n = 1,
    size = I,
    prob = p_IR
  )
  
  # 7. Update compartments
  S_new <- S - N_SI - N_SV
  I_new <- I + N_SI - N_IR
  R_new <- R + N_IR
  V_new <- V + N_SV
  
  # 8. Return new state and incidence
  return(
    list(
      S = S_new,
      I = I_new,
      R = R_new,
      V = V_new,
      incidence = N_SI,
      new_vaccinations = N_SV,
      new_recoveries = N_IR
    )
  )

  #i would do df$columnname[i] and replace it 
}


#defining intital vectors as was in the project
df1 <- initial_populations(1-0.004, 0.004, 0, T = T)
print(head(df1))


for (i in 1:T){
    
}





simulate_markov <- function(beta, gamma, mu, x0, y0, v0, T, N = 56490048) {
  
  df <- initial_populations(
    x0 = x0,
    y0 = y0,
    v0 = v0,
    T = T,
    N = N
  )
  
  for (i in 1:T) {
    
    step <- one_step_markov(
      S = df$S[i],
      I = df$I[i],
      R = df$R[i],
      V = df$V[i],
      beta = beta,
      gamma = gamma,
      mu = mu,
      N = N
    )
    
    df$S[i + 1] <- step$S
    df$I[i + 1] <- step$I
    df$R[i + 1] <- step$R
    df$V[i + 1] <- step$V
    
    df$incidence[i + 1] <- step$incidence
    
    df$x[i + 1] <- step$S / N
    df$y[i + 1] <- step$I / N
    df$r[i + 1] <- step$R / N
    df$v[i + 1] <- step$V / N
  }
  
  return(df)
}

```

defining an N_eff because this should display a little more stochastic variance. useful for visualisation.


```{r}

T <- 1000

#df1 <- simulate_markov(
  #beta = 0.15,
  #gamma = 0.1,
  #mu = 0.001,
  #x0 = 1 - 0.004,
  #y0 = 0.004,
  #v0 = 0,
  #T = T,
  #N = 56490048
#)

#print(head(df1))


#plot(df1$t, df1$incidence / 56490048, type = "l")




library(ggplot2)

# Real UK population used for observed case proportions
N_UK <- 56490048

# Effective population size used inside stochastic Markov model
# Smaller N_eff = more visible stochastic variation
N_eff <- 50000

T <- 1000

set.seed(4)

df1 <- simulate_markov(
    
  beta = best_beta,
  gamma = 0.1,
  mu = 0.000,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  T = 46,
  N = N_eff
)

omicron_data <- read.csv("/Users/willlodder/Desktop/Math245/omicron.csv")
omicron_data$date <- as.Date(omicron_data$date)

# Remove initial model row because incidence[1] is artificial
model_incidence <- df1[-1, c("t", "incidence")]

n <- min(nrow(omicron_data), nrow(model_incidence))

comparison_df <- data.frame(
  date = omicron_data$date[1:n],
  observed_cases = omicron_data$cases[1:n],
  observed_prop = omicron_data$cases[1:n] / N_UK,
  model_incidence = model_incidence$incidence[1:n],
  model_prop = model_incidence$incidence[1:n] / N_eff
)

p1 <- ggplot(comparison_df, aes(x = date)) +
  geom_line(aes(y = observed_prop, linetype = "Observed daily cases / N")) +
  geom_line(aes(y = model_prop, linetype = "Markov model incidence / N_eff")) +
  labs(
    x = "Date",
    y = "Daily incidence proportion",
    title = "Observed Omicron Daily Cases vs Markov Model Incidence",
    linetype = "Series"
  ) +
  theme_minimal()




print(p1)


```


This appears way off. this is because we are using the value beta_best which calculates daily indicence as the number of new infected individuals - number of new recovered individuals, which is not used in practice. will now make a code block to fit a better parameter for beta for our markov model

```{r}

# ------------------------------------------------------------
# Fit beta using the correct Markov-style incidence definition
# incidence = new infections / N
# ------------------------------------------------------------

markov_expected_incidence <- function(beta, gamma, mu, x0, y0, v0, T) {
  
  # Store proportions, not integer counts
  x <- numeric(T + 1)
  y <- numeric(T + 1)
  r <- numeric(T + 1)
  v <- numeric(T + 1)
  incidence <- numeric(T + 1)
  
  # Initial conditions
  x[1] <- x0
  y[1] <- y0
  v[1] <- v0
  r[1] <- 1 - x0 - y0 - v0
  incidence[1] <- 0
  
  for (i in 1:T) {
    
    # Markov hazards for one susceptible individual
    lambda_inf <- beta * y[i]
    lambda_vac <- mu
    lambda_total <- lambda_inf + lambda_vac
    
    # One-day probabilities for S -> I and S -> V
    if (lambda_total > 0) {
      p_leave_S <- 1 - exp(-lambda_total)
      p_SI <- (lambda_inf / lambda_total) * p_leave_S
      p_SV <- (lambda_vac / lambda_total) * p_leave_S
    } else {
      p_SI <- 0
      p_SV <- 0
    }
    
    # One-day probability for I -> R
    p_IR <- 1 - exp(-gamma)
    
    # Expected proportions moving between compartments
    new_infections <- x[i] * p_SI
    new_vaccinations <- x[i] * p_SV
    new_recoveries <- y[i] * p_IR
    
    # Update proportions
    x[i + 1] <- x[i] - new_infections - new_vaccinations
    y[i + 1] <- y[i] + new_infections - new_recoveries
    r[i + 1] <- r[i] + new_recoveries
    v[i + 1] <- v[i] + new_vaccinations
    
    # Correct incidence: new infections, not y[i+1] - y[i]
    incidence[i + 1] <- new_infections
  }
  
  data.frame(
    t = 0:T,
    x = x,
    y = y,
    r = r,
    v = v,
    incidence = incidence
  )
}


markov_error_beta <- function(beta, data, gamma, mu, x0, y0, v0, N_UK) {
  
  # Observed daily case proportion
  observed_incidence <- data$cases / N_UK
  
  # We simulate one interval per observed daily case
  T_fit <- length(observed_incidence)
  
  model <- markov_expected_incidence(
    beta = beta,
    gamma = gamma,
    mu = mu,
    x0 = x0,
    y0 = y0,
    v0 = v0,
    T = T_fit
  )
  
  # Remove incidence[1], because it is artificial
  model_incidence <- model$incidence[-1]
  
  # Safety check
  n <- min(length(observed_incidence), length(model_incidence))
  
  sum((model_incidence[1:n] - observed_incidence[1:n])^2)
}


# ------------------------------------------------------------
# Actually find beta_best_markov
# ------------------------------------------------------------

N_UK <- 56490048

fit_markov <- optimize(
  f = markov_error_beta,
  interval = c(0.01, 0.5),
  data = omicron_data,
  gamma = 0.1,
  mu = 0,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  N_UK = N_UK
)

beta_best_markov <- fit_markov$minimum
markov_min_error <- fit_markov$objective

print(beta_best_markov)
print(markov_min_error)

```


We now use this corrected beta value and plot a single path using this. 
```{r}

df2 <- simulate_markov(
    
  beta = beta_best_markov,
  gamma = 0.1,
  mu = 0.000,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  T = 46,
  N = N_eff
)

omicron_data <- read.csv("/Users/willlodder/Desktop/Math245/omicron.csv")
omicron_data$date <- as.Date(omicron_data$date)

# Remove initial model row because incidence[1] is artificial
model_incidence_2 <- df2[-1, c("t", "incidence")]

n <- min(nrow(omicron_data), nrow(model_incidence_2))

comparison_df_2 <- data.frame(
  date = omicron_data$date[1:n],
  observed_cases = omicron_data$cases[1:n],
  observed_prop = omicron_data$cases[1:n] / N_UK,
  model_incidence = model_incidence_2$incidence[1:n],
  model_prop = model_incidence_2$incidence[1:n] / N_eff
)

p2 <- ggplot(comparison_df_2, aes(x = date)) +
  geom_line(aes(y = observed_prop, linetype = "Observed daily cases / N")) +
  geom_line(aes(y = model_prop, linetype = "Markov model incidence / N_eff")) +
  labs(
    x = "Date",
    y = "Daily incidence proportion",
    title = "Observed Omicron Daily Cases vs Markov Model Incidence",
    linetype = "Series"
  ) +
  theme_minimal()




print(p2)



```



We now want to find confidence bands, by taking M simulations and taking quantiles. 

```{r}

# ------------------------------------------------------------
# Extension: simulate many Markov paths and form uncertainty bands
# ------------------------------------------------------------

simulate_many_markov <- function(M, beta, gamma, mu, x0, y0, v0, T, N_eff) {
  
  simulations <- vector("list", M)
  
  for (m in 1:M) {
    
    sim <- simulate_markov(
      beta = beta,
      gamma = gamma,
      mu = mu,
      x0 = x0,
      y0 = y0,
      v0 = v0,
      T = T,
      N = N_eff
    )
    
    # Remove first artificial incidence value
    sim <- sim[-1, ]
    
    simulations[[m]] <- data.frame(
      simulation = m,
      t = sim$t,
      incidence_prop = sim$incidence / N_eff
    )
  }
  
  do.call(rbind, simulations)
}


set.seed(123)

M <- 5000
T_fit <- nrow(omicron_data)

many_sims <- simulate_many_markov(
  M = M,
  beta = beta_best_markov,
  gamma = 0.1,
  mu = 0,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  T = T_fit,
  N_eff = N_eff
)
# ------------------------------------------------------------
# Summarise simulations at each time point
# Safer version: avoids quantile name issues
# ------------------------------------------------------------

simulation_summary <- do.call(
  rbind,
  lapply(split(many_sims$incidence_prop, many_sims$t), function(z) {
    data.frame(
      mean = mean(z),
      lower = as.numeric(quantile(z, 0.025)),
      upper = as.numeric(quantile(z, 0.975))
    )
  })
)

simulation_summary$t <- as.numeric(rownames(simulation_summary))
rownames(simulation_summary) <- NULL

# Reorder columns nicely
simulation_summary <- simulation_summary[, c("t", "mean", "lower", "upper")]

# Add dates and observed incidence
n <- min(nrow(omicron_data), nrow(simulation_summary))

plot_df <- data.frame(
  date = as.Date(omicron_data$date[1:n]),
  observed_prop = omicron_data$cases[1:n] / N_UK,
  mean = simulation_summary$mean[1:n],
  lower = simulation_summary$lower[1:n],
  upper = simulation_summary$upper[1:n]
)


p_many <- ggplot(plot_df, aes(x = date)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.25) +
  geom_line(aes(y = mean, linetype = "Mean Markov incidence")) +
  geom_line(aes(y = observed_prop, linetype = "Observed daily cases / N")) +
  labs(
    x = "Date",
    y = "Daily incidence proportion",
    title = "Observed Omicron Cases vs Markov Simulation Envelope",
    linetype = "Series"
  ) +
  theme_minimal()

print(p_many)



```



























```{r}

# ------------------------------------------------------------
# Animated SIRV Markov compartment flow
# S -> I, I -> R, S -> V
# ------------------------------------------------------------

# install.packages(c("gganimate", "gifski"))

library(ggplot2)
library(gganimate)
library(gifski)
library(grid)

set.seed(245)

# ------------------------------------------------------------
# Version of your Markov simulator which stores daily flows
# ------------------------------------------------------------

simulate_markov_flow <- function(beta, gamma, mu, x0, y0, v0, T, N = 50000) {
  
  df <- initial_populations(
    x0 = x0,
    y0 = y0,
    v0 = v0,
    T = T,
    N = N
  )
  
  df$new_infections <- integer(T + 1)
  df$new_vaccinations <- integer(T + 1)
  df$new_recoveries <- integer(T + 1)
  
  for (i in 1:T) {
    
    step <- one_step_markov(
      S = df$S[i],
      I = df$I[i],
      R = df$R[i],
      V = df$V[i],
      beta = beta,
      gamma = gamma,
      mu = mu,
      N = N
    )
    
    df$S[i + 1] <- step$S
    df$I[i + 1] <- step$I
    df$R[i + 1] <- step$R
    df$V[i + 1] <- step$V
    
    df$incidence[i + 1] <- step$incidence
    df$new_infections[i + 1] <- step$incidence
    df$new_vaccinations[i + 1] <- step$new_vaccinations
    df$new_recoveries[i + 1] <- step$new_recoveries
    
    df$x[i + 1] <- step$S / N
    df$y[i + 1] <- step$I / N
    df$r[i + 1] <- step$R / N
    df$v[i + 1] <- step$V / N
  }
  
  return(df)
}

# ------------------------------------------------------------
# Parameters for the animation
# ------------------------------------------------------------

N_eff_anim <- 50000
T_anim <- 80

beta_anim <- if (exists("beta_best_markov")) beta_best_markov else 0.1357475
gamma_anim <- 0.1
mu_anim <- 0.002

flow_sim <- simulate_markov_flow(
  beta = beta_anim,
  gamma = gamma_anim,
  mu = mu_anim,
  x0 = 1 - 0.004,
  y0 = 0.004,
  v0 = 0,
  T = T_anim,
  N = N_eff_anim
)

# ------------------------------------------------------------
# Layout of the four compartments
# ------------------------------------------------------------

box_w <- 1.35
box_h <- 0.85

layout_df <- data.frame(
  compartment = c("S", "I", "R", "V"),
  full_name = c("Susceptible", "Infectious", "Removed", "Vaccinated"),
  cx = c(0, 2.7, 5.4, 0),
  cy = c(0.6, 0.6, 0.6, -1.45)
)

# ------------------------------------------------------------
# Build animated compartment boxes
# ------------------------------------------------------------

substeps <- 10
box_frames <- list()
k <- 1

for (d in 1:T_anim) {
  
  previous <- flow_sim[d, ]
  next_state <- flow_sim[d + 1, ]
  
  for (j in 0:(substeps - 1)) {
    
    phase <- j / (substeps - 1)
    frame_id <- (d - 1) * substeps + j + 1
    
    counts <- c(
      S = (1 - phase) * previous$S + phase * next_state$S,
      I = (1 - phase) * previous$I + phase * next_state$I,
      R = (1 - phase) * previous$R + phase * next_state$R,
      V = (1 - phase) * previous$V + phase * next_state$V
    )
    
    for (comp in names(counts)) {
      
      this_layout <- layout_df[layout_df$compartment == comp, ]
      prop <- counts[comp] / N_eff_anim
      
      box_frames[[k]] <- data.frame(
        frame = frame_id,
        day = d - 1,
        compartment = comp,
        full_name = this_layout$full_name,
        count = counts[comp],
        prop = prop,
        cx = this_layout$cx,
        cy = this_layout$cy,
        xmin = this_layout$cx - box_w / 2,
        xmax = this_layout$cx + box_w / 2,
        ymin = this_layout$cy - box_h / 2,
        ymax = this_layout$cy + box_h / 2,
        bar_ymax = this_layout$cy - box_h / 2 + prop * box_h,
        label = paste0(
          this_layout$full_name,
          "\n",
          round(100 * prop, 2),
          "%"
        )
      )
      
      k <- k + 1
    }
  }
}

box_frames <- do.call(rbind, box_frames)

# ------------------------------------------------------------
# Day label
# ------------------------------------------------------------

day_labels <- unique(box_frames[, c("frame", "day")])

day_labels$label <- paste0(
  "Day ", day_labels$day,
  "     ",
  expression(beta)
)

day_labels$main_label <- paste0(
  "Day ", day_labels$day,
  "     beta = ", round(beta_anim, 4),
  "     gamma = ", gamma_anim,
  "     mu = ", mu_anim
)

# ------------------------------------------------------------
# Static arrows between compartments
# ------------------------------------------------------------

arrow_df <- data.frame(
  transition = c("S to I", "I to R", "S to V"),
  x = c(
    layout_df$cx[layout_df$compartment == "S"] + box_w / 2,
    layout_df$cx[layout_df$compartment == "I"] + box_w / 2,
    layout_df$cx[layout_df$compartment == "S"]
  ),
  y = c(
    layout_df$cy[layout_df$compartment == "S"],
    layout_df$cy[layout_df$compartment == "I"],
    layout_df$cy[layout_df$compartment == "S"] - box_h / 2
  ),
  xend = c(
    layout_df$cx[layout_df$compartment == "I"] - box_w / 2,
    layout_df$cx[layout_df$compartment == "R"] - box_w / 2,
    layout_df$cx[layout_df$compartment == "V"]
  ),
  yend = c(
    layout_df$cy[layout_df$compartment == "I"],
    layout_df$cy[layout_df$compartment == "R"],
    layout_df$cy[layout_df$compartment == "V"] + box_h / 2
  ),
  arrow_label = c(
    "infection",
    "recovery",
    "vaccination"
  ),
  label_x = c(1.35, 4.05, -0.72),
  label_y = c(0.88, 0.88, -0.42)
)

# ------------------------------------------------------------
# Helper function for moving particles along an arrow
# ------------------------------------------------------------

make_particles <- function(day, count, transition_name, x0, y0, x1, y1,
                           max_count, max_particles = 35) {
  
  if (count <= 0 || max_count <= 0) {
    return(NULL)
  }
  
  n_particles <- max(
    1,
    ceiling(max_particles * sqrt(count / max_count))
  )
  
  dx <- x1 - x0
  dy <- y1 - y0
  len <- sqrt(dx^2 + dy^2)
  
  px <- -dy / len
  py <- dx / len
  
  out <- list()
  k <- 1
  
  for (p in 1:n_particles) {
    
    offset <- runif(1, -0.25, 0.25)
    side_jitter <- rnorm(1, mean = 0, sd = 0.035)
    
    for (j in 0:(substeps - 1)) {
      
      raw_phase <- j / (substeps - 1) + offset
      phase <- min(max(raw_phase, 0), 1)
      
      frame_id <- (day - 1) * substeps + j + 1
      
      out[[k]] <- data.frame(
        frame = frame_id,
        particle_id = paste(transition_name, day, p, sep = "_"),
        transition = transition_name,
        x = x0 + phase * dx + side_jitter * px,
        y = y0 + phase * dy + side_jitter * py
      )
      
      k <- k + 1
    }
  }
  
  do.call(rbind, out)
}

# ------------------------------------------------------------
# Create all moving particles
# ------------------------------------------------------------

max_flow_count <- max(
  flow_sim$new_infections,
  flow_sim$new_recoveries,
  flow_sim$new_vaccinations
)

particle_list <- list()
k <- 1

for (d in 1:T_anim) {
  
  # Transitions from day d - 1 to day d are stored in row d + 1
  row_now <- flow_sim[d + 1, ]
  
  particle_list[[k]] <- make_particles(
    day = d,
    count = row_now$new_infections,
    transition_name = "S to I",
    x0 = arrow_df$x[arrow_df$transition == "S to I"],
    y0 = arrow_df$y[arrow_df$transition == "S to I"],
    x1 = arrow_df$xend[arrow_df$transition == "S to I"],
    y1 = arrow_df$yend[arrow_df$transition == "S to I"],
    max_count = max_flow_count
  )
  k <- k + 1
  
  particle_list[[k]] <- make_particles(
    day = d,
    count = row_now$new_recoveries,
    transition_name = "I to R",
    x0 = arrow_df$x[arrow_df$transition == "I to R"],
    y0 = arrow_df$y[arrow_df$transition == "I to R"],
    x1 = arrow_df$xend[arrow_df$transition == "I to R"],
    y1 = arrow_df$yend[arrow_df$transition == "I to R"],
    max_count = max_flow_count
  )
  k <- k + 1
  
  particle_list[[k]] <- make_particles(
    day = d,
    count = row_now$new_vaccinations,
    transition_name = "S to V",
    x0 = arrow_df$x[arrow_df$transition == "S to V"],
    y0 = arrow_df$y[arrow_df$transition == "S to V"],
    x1 = arrow_df$xend[arrow_df$transition == "S to V"],
    y1 = arrow_df$yend[arrow_df$transition == "S to V"],
    max_count = max_flow_count
  )
  k <- k + 1
}

particles_df <- do.call(rbind, particle_list)

# ------------------------------------------------------------
# Plot and animate
# ------------------------------------------------------------

p_flow <- ggplot() +
  
  geom_segment(
    data = arrow_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.22, "cm")),
    linewidth = 0.8,
    alpha = 0.75
  ) +
  
  geom_text(
    data = arrow_df,
    aes(x = label_x, y = label_y, label = arrow_label),
    size = 3.5
  ) +
  
  geom_rect(
    data = box_frames,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      ymax = ymax,
      fill = compartment
    ),
    colour = "black",
    linewidth = 0.7,
    alpha = 0.20
  ) +
  
  geom_rect(
    data = box_frames,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      ymax = bar_ymax,
      fill = compartment
    ),
    alpha = 0.55
  ) +
  
  geom_text(
    data = box_frames,
    aes(x = cx, y = cy, label = label),
    size = 4,
    fontface = "bold",
    lineheight = 0.95
  ) +
  
  geom_point(
    data = particles_df,
    aes(x = x, y = y, colour = transition, group = particle_id),
    size = 2.2,
    alpha = 0.9
  ) +
  
  geom_text(
    data = day_labels,
    aes(x = 2.7, y = 1.45, label = main_label),
    size = 4.5,
    fontface = "bold"
  ) +
  
  scale_fill_manual(
    values = c(
      S = "#4C78A8",
      I = "#E45756",
      R = "#72B7B2",
      V = "#54A24B"
    ),
    name = "Compartment"
  ) +
  
  scale_colour_manual(
    values = c(
      "S to I" = "#E45756",
      "I to R" = "#72B7B2",
      "S to V" = "#54A24B"
    ),
    name = "Transition"
  ) +
  
  coord_equal(
    xlim = c(-1.25, 6.25),
    ylim = c(-2.15, 1.7),
    expand = FALSE
  ) +
  
  labs(
    title = "Animated SIRV Markov Compartment Flow",
    subtitle = "Particles show daily stochastic transitions between compartments"
  ) +
  
  theme_void() +
  
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "bottom"
  ) +
  
  transition_manual(frame)

flow_animation <- animate(
  p_flow,
  nframes = length(unique(box_frames$frame)),
  fps = 20,
  width = 950,
  height = 560,
  renderer = gifski_renderer()
)

anim_save("markov_compartment_flow.gif", flow_animation)

```
